``Interestingness'' in Mathematical Theory Exploration

### 2019-11-29 PhDThesis/bucketing.tex
Finish off bucketing

\begin{enumerate}
\item Define the \emph{signature selection problem} for theory exploration.
\item Analyse the effect of signature selection on the quantity of desirable
  properties attainable by theory exploration on the TEB problem set.
\item Identify the effect of ``toxic'' definitions on system performance, as
  measured by the TEB.
\item Recommend improvements for theory exploration systems, to avoid or
  mitigate the problem of toxic definitions.
\item Demonstrate how the signature selection problem is distinct from the
  well-known \emph{clustering problem} in machine learning, and how clustering
  algorithms cannot be directly applied to signature selection.
\end{enumerate}

@@

\begin{figure}
  \scalebox{0.69}{\input{images/proportionsall.pgf}}
  \scalebox{0.69}{\input{images/proportionsnontoxic.pgf}}
  \caption[]{Each bar is a definition from our Theory Exploration Benchmark,
    showing the proportion of signatures they appeared in which were
    successfully explored (green) compared to those which timed out (red).
    Definitions are sorted from most successful to least. The left graph shows
    all \input{text/proportions_graph_count_all.tex} TEB definitions explored, with five ``toxic'' definitions which always
    failed. with s ratio of successes to failures of the runs they appeared in. The first graph contains all TIP definitions, showing
    ``toxic'' definitions which always failed. The second graph only contains
    runs without any toxic definitions.}
  \label{fig:proportions}
\end{figure}

@@

\begin{figure}
  % \input{images/bounds.pgf}
  \includegraphics[width=\textwidth]{images/bounds.png}
  \caption{The average quality (proportion of ground-truth properties available)
    when splitting signatures sampled using the Theory Exploration Benchmark
    into subsets. Each plot shows division into a different number of subsets.
    The shaded area is bounded by the optimal and pessimal methods, which can
    only produce $N$ subsets for signatures of size $\geq N$. Grey lines show
    pseudorandom selection, for comparison; these extend into smaller signature
    sizes since empty subsets are permitted.}
  \label{fig:bounds}
\end{figure}

@@

\begin{figure}
  % \input{images/bucketed}
  \includegraphics[width=\textwidth]{images/bucketed.png}
  \caption{Quality evaluation of bucketing via recurrent clustering. Each plot
    uses a fixed number of buckets (5, 10, 15 or 20) and shows the number of
    extra theorems available when clusters are used as buckets, compared to a
    pseudorandom control. The difference is not statistically significant. There
    is a slight upward trend for larger sizes, but this is relatively small
    compared to the corresponding increase in total theorems available.}
  \label{fig:bucketing}
\end{figure}

The signature selection problem is superficially similar to the
\emph{clustering} problem in machine learning: grouping data points based on
their similarity across multiple dimensions. To investigate this potential
relationship we created a signature selection method using the \emph{recurrent
  clustering} algorithm for Haskell code described in
\S~\ref{sec:recurrentclustering}: this flattens syntax trees into vectors,
substitutes fixed numbers for keywords and recurses to determine suitable
numbers for referenced expressions. These vectors are clustered based on their
euclidean distance, using a \emph{k-means} algorithm, and these clusters are
chosen as our buckets.

The results of this method are shown in Figure~\ref{fig:bucketing}, which plots
the difference between bucketing based on this clustering approach, compared to
simply choosing buckets pseudorandomly (as a control). The results of clustering
are statistically indistinguishable from random.\footnote{Wilcoxon's signed-rank
  test finds no significant difference for most of the plotted points; whilst
  not a rigorous comparison (hence its ommission), this is nevertheless a clear
  indication of their similarity.}

\iffalse
TODO: Alison: Does this translate into a recommendation (e.g. not to
use clustering)? There should be some discussion of this - if not here then
point to another chapter where you discuss these results

TODO: Alison: Needs a summary section
\fi

>>>>

\begin{enumerate}
\item Define the \emph{signature selection problem} for theory exploration.
\item Identify the effect of ``toxic'' definitions on system performance, as
  measured by the TEB.
\item Recommend improvements for theory exploration systems, such as
  resource-bounding of individual evaluations, to avoid or mitigate the
  problem of toxic definitions.
\item Analyse the effect of signature selection on the quality of discovered
  results, by measuring how many desirable properties of a benchmark problem
  (TEB) remain discoverable after splitting into multiple small signatures.
\item Demonstrate how the signature selection problem is distinct from the
  well-known \emph{clustering problem} in machine learning, and how clustering
  algorithms cannot be directly applied to signature selection.
\end{enumerate}

@@

\begin{figure}
  \scalebox{0.69}{\input{images/proportionsall.pgf}}
  \scalebox{0.69}{\input{images/proportionsnontoxic.pgf}}
  \caption[]{Each vertical bar is a definition from our Theory Exploration
    Benchmark (TEB), showing the proportion of signatures they appeared in which
    were successfully explored (green) compared to those which timed out (red).
    Definitions are sorted from most successful to least. The left graph shows
    all \input{text/proportions_graph_count_all.tex} TEB definitions explored,
    where five ``toxic'' definitions always failed. In comparison, the right
    graph ignores signatures containing those five definitions; the much higher
    proportion of success highlights the disproportionate impact a few toxic
    definitions can have.}
  \label{fig:proportions}
\end{figure}

@@

\begin{figure}
  % \input{images/bounds.pgf}
  \includegraphics[width=\textwidth]{images/bounds.png}
  \caption{The average quality (proportion of ground-truth properties available)
    when selecting subsets of definitions sampled from the Theory Exploration
    Benchmark. Each plot shows division into a different number of subsets.
    The shaded area is bounded by the optimal and pessimal methods, whose
    constraints prevent selecting $N$ signatures if there are fewer than $N$
    definitions to pick from. Grey lines show pseudorandom selection for
    comparison; these are unconstrained, so extend into smaller signature
    sizes.}
  \label{fig:bounds}
\end{figure}

@@

\begin{figure}
  % \input{images/bucketed}
  \includegraphics[width=\textwidth]{images/bucketed.png}
  \caption{Evaluating the quality of selecting signatures using recurrent
    clustering. Each plot uses a fixed number of subsets (5, 10, 15 or 20) and
    shows the number of extra theorems available when clusters are used as these
    subsets, compared to a pseudorandom control. The difference is not
    statistically significant. There is a slight upward trend for larger sizes,
    but this is relatively small compared to the corresponding increase in total
    theorems available.}
  \label{fig:bucketing}
\end{figure}

The signature selection problem is superficially similar to the
\emph{clustering} problem in machine learning: grouping data points based on
their similarity across multiple dimensions. To investigate this potential
relationship we created a signature selection method using the \emph{recurrent
  clustering} algorithm for Haskell code described in
\S~\ref{sec:recurrentclustering}: this flattens syntax trees into vectors,
substitutes fixed numbers for keywords and recurses to determine suitable
numbers for referenced expressions. These vectors are clustered based on their
euclidean distance, using a \emph{k-means} algorithm, and these clusters are
chosen as the subsets to explore separately.

The results of this method are shown in Figure~\ref{fig:bucketing}, which plots
the difference between selecting clusters of definitions, compared to simply
choosing pseudorandomly (as a control). The results of clustering are
statistically indistinguishable from random.\footnote{Wilcoxon's signed-rank
  test finds no significant difference for most of the plotted points; whilst
  not a rigorous comparison (hence its ommission), this is nevertheless a clear
  indication of their similarity.} Despite this, two trends are apparent: a
slight increase in clustering's (relative) quality as the sample size increases,
and a decrease in variance as the number of subsets is increased. The variance
may be explained by the larger number of permutations available when using more
subsets, which decreases the chance of finding an outlier (either good \emph{or}
bad). The absolute increase in theorems found by clustering for large sizes, if
real, is dwarfed by the corresponding increase in the ground truth (the total
number of theorems available, without selecting subsets); the quality difference
between these selection methods as a \emph{proportion} of the ground truth
actually \emph{decreases} for larger sizes.

\section{Summary}

We have identified the \emph{signature selection problem} for theory
exploration, which narrows the available definitions (e.g. from a software
library) down to a subset fast enough to explore. This must currently be solved
by the user, which imposes a knowledge burden (reducing TE's utility as an aid
to learning) and limits how novel or surprising the discovered properties can
be.

We have demonstrated the effect of signature size (and hence the importance of
selection) on the time taken to explore, in particular noting the ``heavy tail''
of explorations timing out, which increases linearly with signature size. We
have attributed this to ``toxic'' definitions, whose presence in a signature
will exhaust the allocated time and/or memory. To mitigate this problem, and
hence reduce the severity of signature selection, we recommend a more
fine-grained approach to resource management and/or scheduling: by limiting the
resources of individual tests, evaluation of toxic definitions can be terminated
without having to abort the whole exploration run.

We have shown the impact of signature selection on the quality of discoverable
properties, by sampling sets of definitions from the Theory Exploration
Benchmark and comparing how many of their associated theorems become impossible
to discover (due to their definitions being separated) as these sets are
subdivided by different selection methods. Optimal selection (calculated with
knowledge of the benchmark theorems) retains a large proportion of the desired
output, showing the potential to maintain high quality. Pseudorandom selection
is found to be closer to pessimal, and selecting based on similarity (determined
via an unsupervised machine learning algorithm) performs no different to random.

Automated signature selection is hence an important problem, requiring further
research to establish if, and how, the potential benefits can be reaped.
